Evaluation of +S2 sea ice forecast and some other better title

### 2025-03-07 analysis/paper/access-ice.qmd
Changes antarctica map and other stuff

The +S2 hindcast used in this study runs for the period 1981--2018.
Each forecast consists of 9 ensemble members, 3 of which run for 279 days and 6 for 42 days.
Anomalies are be taken with respect to the 1981--2011 climatology computed from the daily +S2 reanalysis and smoothed with a 11 day running mean.

The +S1 hindcast runs for the period 1990--2012, with 7-member forecasts integrated for 217 days. 
No reanalysis is available so we are not able to compute a climatology for this forecast, therefore we don't show metrics based on anomalies.

@@

```{r extent_climatology}
extent_climatology <- extent_daily |>  
  copy() |>
  _[aice == 0, aice := NA] |> 
  _[!(month(time) == 2 & mday(time) == 29)] |> 
  _[, time := update(time, year = 2001)] |> 
  _[, median_ci(aice), by = .(time, dataset)]

```{r area_daily}
area_daily <- datasets |>
  lapply(area) |>
  lapply(\(x) ReadNetCDF(x, "aice")) |>
  rbindlist(idcol = "dataset") |>
  _[, let(lon = NULL, lat = NULL)]
```

```{r area_climatology}
area_climatology <- area_daily |>  
  copy() |>
  _[aice == 0, aice := NA] |> 
  _[!(month(time) == 2 & mday(time) == 29)] |> 
  _[, time := update(time, year = 2001)] |> 
  _[, median_ci(aice), by = .(time, dataset)]

@@

```{r fig-hindcast-extent}
#| fig-cap: Median sea ice extent for al hindcasts initialised the first of the month for +S2 and +S1 in colours representing the start month. In black, the median sea ice extent of +CDR. 
hindcast_extent |> 
  _[, median_ci(aice), by = .(forecast_time = update(forecast_time, year = 2001),
                              lag = as.numeric(as.Date(time) - forecast_time),
                              model)] |>
  _[, time := forecast_time + lag] |>
  _[, time2 := update(time, year = 2001)] |>
  _[, month := lubridate::month(forecast_time, label = TRUE)] |>
  # _[]
  ggplot(aes(time2, mid)) +
  geom_line(data = extent_climatology[dataset == which_dataset], 
            aes(x = as.Date(time),
                group = dataset)) + 
  geom_line(aes(color = month, group = interaction(forecast_time, time != time2))) +

@@

  scale_x_date(NULL, date_breaks = "1 month", 
               date_labels = "%b", expand = c(0, 0)) +
  scale_y_continuous(NULL, labels = labels_extent) +
  scale_color_manual(values = cetcolor::cet_pal(12, name = "c2"), 
                     aesthetics = c("fill", "color"), guide = "none") +
  facet_wrap(~ model, ncol = 1) +
  tag_facets() 
```

@fig-hindcast-extent a shows the seasonal cycle of Sea Ice Extent for the +S2 hindcast and +CDR +S2 shows a severe low extent bias, especially in the late summer-early autumn.
This is due primarily to faster melt during a longer melt season between January and March and slower growth during March and April. 
This is then balanced with faster growth between May and July ([Fig. @fig-mean-growth]).
Many sea ice models exhibit this systematic underestimation during the sea ice minimum and early freezing season [@massonnet2023] and could indicate problems in the representation of thermodynamics in the model [@zampieri2019].

+S1 has a smaller low bias than +S2, especially at low lags ([Fig. @fig-hindcast-extent b]) even though the typical growth rates are very similar between both models ([Fig. @fig-mean-growth]).
At larger lags, +S1's bias in summer and autumn is very similar to +S2's.
This suggests that this lower sea ice state is closer to the models' equilibrium, indicating that it is an issue with model formulation that was being corrected by the data assimilation system in +S1 and not in +S2.

At large lags, sea ice extent loses most of the initial condition memory and reverts to the model's preferred equilibrium state.
Therefore we can estimate the latter using the hindcasts with the largest possible lag, which is shown in triangles in @fig-hindcast-extent for the same dates as the initial conditions.
The difference between the two is the effect of the data assimilation.

The equilibrium of +S1 and +S2 is very similar (comparing the triangles in each panel in [Fig. @fig-hindcast-extent]), owing to them having the same model formulation.
From June to October, in +S2 circles are closer to observations than to the triangles, indicating that the information from the ocean and atmosphere data assimilation is getting to sea ice and improving the initial conditions.
The rest of the year, there is little if any difference between circles and triangles in +S2, indicating that almost no data assimilation is taking place and the sea ice component of the model is virtually free-running.
In +S1, on the other hand, circles are very close to observations year-round.

@@

dif_hindcast <- hindcast_extent |> 
  copy() |> 
  _[order(time)] |> 
  _[, dif := c(NA, diff(aice)/diff(as.numeric(as.Date(time)))), by = .(model, forecast_time, member)] |> 
  _[, median_ci(dif), by = .(model, time = update(time, year = 2000))]

extent_daily |> 
  _[order(time)] |> 
  _[, dif := c(NA, diff(aice)/diff(as.numeric(as.Date(time)))), by = .(dataset)] |> 
  _[, median_ci(dif), by = .(dataset, time = update(time, year = 2000))] |> 
  _[dataset == which_dataset] |> 
  ggplot(aes(time, mid)) +
  geom_hline(yintercept = 0, color = "gray50", linewidth = 0.2) +
  geom_smooth(aes(group = dataset, color = dataset),
              span = 30/366, linewidth = 0.5, se = FALSE, n = 366)  +
  geom_smooth(data = dif_hindcast,
              aes(color = model), span = 30/366, linewidth = 0.5, se = FALSE, n = 366) +
  scale_x_datetime(NULL, date_breaks = "1 month", 
                   date_labels = "%b", expand = c(0, 0)) +
  scale_y_continuous(NULL, labels = \(x) paste0(labels_extent(x), "/day")) +
  scale_color_models
```

@@

hindcast_clim |> 
  # dcast(time + xgrid + ygrid ~ model, value.var = "aice") |> 
  # setnames(which_dataset, "obs") |>
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = aice - obs, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised(paste0("Bias with respect to ",
                                          labels_models[which_dataset]),
                                   low = scales::muted("red"),
                                   high = scales::muted("blue")) +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ month) +
  wide_legend +
  tag_facets()
```

@@

errors |> 
  _[measure == "rmse"] |>
  _[, time2 := update(time, year = 2000)] |> 
  # merge(clim_std) |>
  _[obs_dataset ==  which_dataset] |>
  _[, median_ci(value),
    by = .(lag, version, measure, month(time_forecast), obs_dataset)] |>
  _[version == "persistence", version := obs_dataset] |> 
  ggplot(aes(lag, mid)) +

@@

  geom_ribbon(aes(ymin = low, 
                  ymax = high, 
                  color = version, 
                  fill = version,
                  group = interaction(obs_dataset, version)),
              alpha = 0.1) +
  geom_line(aes(color = version, group = interaction(obs_dataset, version)), linewidth = 1) +

  # geom_line(aes(y = clim_std)) +
  scale_x_continuous(breaks = NULL, expand = c(0, 0)) +
  labs(y = "RMSE", 
       x = "Lag") +
  scale_color_models +
  scale_fill_models +
  #   scale_color_models +
  #   scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month)) 
```

@@

Use climatology

@@

```{r}
topo_lonlat <- rcdo::cdo_topo(grid = "r180x90") |>
  rcdo::cdo_execute(options = c("-f nc")) |>
  metR::ReadNetCDF(c(z = "topo"), subset = list(lat = c(-90, -60))) |> 
  ggperiodic::periodic(lon = c(0, 360)) 
```

@@

rmse_lon_mean |> 
  _[obs_dataset == which_dataset] |> 
  ggperiodic::periodic(lon = c(0, 360)) |> 
  ggplot(aes(lon, lag)) +
  geom_contour_fill(aes(z = dif, fill = after_stat(level)), binwidth = 0.025)  +
  geom_contour_tanaka(aes(z = dif), range = c(0.01, 0.4),  binwidth = 0.025) +
  geom_contour2(data = topo_lonlat,
                aes(lon, scales::rescale(lat, c(1, 150)), z = z),
                breaks = 0) +
  geom_hline(data = month_from_lag(1, 1), aes(yintercept = lag),
             colour = "gray50", linewidth = 0.1) +

@@

  scale_y_continuous(breaks = NULL, expand = c(0, 0)) +
  labs(fill = NULL, 
       y = "Lag") +
  wide_legend +
  facet_wrap(~ month, labeller = labeller(month = labels_month)) +
  theme(panel.background = element_blank(), 
        tagger.panel.tag.background = element_rect(colour = "white", fill = "white")) +
  tag_facets(position = list(x = 0.08, y = 1, hjust = 0, vjust = 1))
```

>>>>

The +S2 hindcast used in this study runs for the period 1981--2018.
Each forecast consists of 9 ensemble members built from three consecutive 3-member forecasts initialised at the first of every month and the two previous days and run from 279 days. 
The +S1 hindcast is built in the same manner and running for the period 1990--2012.

Anomalies for each hindcast are be taken with respect to a forecast-dependent climatology for the period 1990--2012. 
This serves as a first-order correction of model drift.

@@

@@

```{r extent_climatology}
ranges <- hindcast_extent[, .(range = list(range(time))), by = model]

extent_climatology <- purrr::pmap(ranges, \(model, range) {
  extent_daily |>  
    _[time %between% range] |> 
    _[aice == 0, aice := NA] |> 
    _[!(month(time) == 2 & mday(time) == 29)] |> 
    _[, time := update(time, year = 2001)] |> 
    _[, median_ci(aice), by = .(time, dataset)] |> 
    _[, model := model] |> 
    _[]
}) |> 
  rbindlist()
```

```{r range_max_lag}
range_max_lag <- hindcast_extent[mday(time) == 2, .SD[which.max(lag)], by = .(forecast_time, model)] |> 
  _[, .(max = max(lag),
        min = min(lag)),
    by = model]
```

```{r fig-hindcast-extent}
#| fig-cap: Median sea ice extent for al hindcasts initialised the first of the month for +S2 and +S1 in colours representing the start month with the median sea ice extent of +CDR in the left column and the median difference in the right column. The +CDR climatology is computed in the period corresponding to each hindcast. Circles represent the initial conditions at the first of every month and triangles represent the median value at the first of every month forecasted with the largest possible lead time. Colours indicate the initialisation month of the forecast.

hindcast_bias_extent <- hindcast_extent |> 
  _[, median_ci(aice), by = .(forecast_time = update(forecast_time, year = 2001),
                              lag = as.numeric(as.Date(time) - forecast_time),
                              model)] |> 
  _[, time := forecast_time + lag] |>
  _[, time2 := update(time, year = 2001)] |>
  _[, month := lubridate::month(forecast_time, label = TRUE)] |> 
  ggplot(aes(time2, mid)) +
  geom_line(data = extent_climatology[dataset == which_dataset], 
            aes(x = as.Date(time),
                group = dataset)) + 
  geom_line(aes(color = month, group = interaction(forecast_time, time != time2))) +

@@

  scale_x_date(NULL, date_breaks = "1 month", 
               date_labels = "%b", expand = c(0, 0)) +
  scale_y_continuous(NULL, labels = labels_extent) +
  scale_color_manual(values = cetcolor::cet_pal(12, name = "c2"), 
                     aesthetics = c("fill", "color"), guide = "none") +
  facet_wrap(~ model, ncol = 1, labeller = labeller(model = labels_models)) +
  tag_facets()

hindcast_bias_extent_dif <- copy(hindcast_bias_extent)

hindcast_bias_extent_dif$layers[[1]] <- geom_hline(yintercept = 0) 
hindcast_bias_extent_dif$data <- extent_daily[hindcast_extent, on = c("time")] |> 
  _[, median_ci(i.aice - aice), 
    by = .(forecast_time = update(forecast_time, year = 2001),
           lag = as.numeric(as.Date(time) - forecast_time),
           model)] |> 
  _[, time := forecast_time + lag] |>
  _[, time2 := update(time, year = 2001)] |>
  _[, month := lubridate::month(forecast_time, label = TRUE)]

hindcast_bias_extent +
  hindcast_bias_extent_dif

@fig-hindcast-extent a shows the median sea ice extent (left column) and median difference with respect of +CDR for the +S2 and +S1 hindcasts.
Median extent of initial conditions at the first of every month are indicated with circles, while median extent at the same date but forecasted with the largest lead time possible for each model (between `r range_max_lag[model == "S2", min]` and `r range_max_lag[model == "S2", max]` days for +S2 and between `r range_max_lag[model == "S1", min]` and `r range_max_lag[model == "S1", max]` days for +S1). 
At this large lead time, the the information of the initial conditions is essentially lost and the forecast reverts to each model's preferred equilibrium state.

Both models' equilibrium state (triangles) show low bias of sea ice extent, particularly in the late-autumn and winter months.
This is due primarily to faster melt during a longer melt season between January and March and slower growth during March and April. 
This is then partially balanced with faster growth between May and July ([Fig. @fig-mean-growth]).
Many sea ice models exhibit this systematic underestimation during the sea ice minimum and early freezing season [@massonnet2023] and could indicate problems in the representation of thermodynamics in the model [@zampieri2019].
It is not surprising that both forecasting systems converge to a similar climatology because they share the same model configuration.

+S2 initial conditions (circles) also have a low extent bias, especially in the late summer-early autumn, while +S1 initial conditions are very close to observations. 
The faster growth between May and July reduces the low bias in +S2 for forecasts initialised from April to June, but it adds a high bias in +S1.

The difference between the initial conditions and the model preferred state can be attributed to the effect of data assimilation, which in +S2 is due solely to atmospheric and oceanic data assimilation. 
From June to October, in +S2 circles are closer to observations than to the triangles, indicating that the information from the ocean and atmosphere data assimilation is affecting sea ice and improving the initial conditions.
The rest of the year there is little if any difference between circles and triangles in +S2, indicating that almost no data assimilation is taking place and the sea ice component of the model is virtually free-running.

```{r extent-delta-compute, fig.height=11.99/cm, fig.width=24.09/cm}
dif_hindcast <- hindcast_extent |> 
  copy() |> 
  _[order(time)] |> 
  _[, dif := c(NA, diff(aice)/diff(as.numeric(as.Date(time)))), by = .(model, forecast_time, member)] |> 
  _[, median_ci(dif), by = .(model, time = update(time, year = 2000))]

dif_hindcast <- rbind(
  dif_hindcast |> 
    copy() |> 
    _[, time := update(time, year = 1999)],
  dif_hindcast,
  dif_hindcast |> 
    copy() |> 
    _[, time := update(time, year = 2001)]
)

dif_extent <- extent_daily |> 
  _[order(time)] |> 
  _[, dif := c(NA, diff(aice)/diff(as.numeric(as.Date(time)))), by = .(dataset)] |> 
  _[, median_ci(dif), by = .(dataset, time = update(time, year = 2000))] |> 
  _[dataset == which_dataset]

dif_extent <- rbind(
  dif_extent |> 
    copy() |> 
    _[, time := update(time, year = 1999)],
  dif_extent,
  dif_extent |> 
    copy() |> 
    _[, time := update(time, year = 2001)]
)
```

@@

n = 366*3
dif_extent |> 
  ggplot(aes(time, mid)) +
  geom_hline(yintercept = 0, color = "gray50", linewidth = 0.2) +
  geom_smooth(aes(group = dataset, color = dataset), 
              method = "loess",
              span = 30/n, linewidth = 0.5, se = FALSE, n = n)  +
  # geom_line() +
  geom_smooth(data = dif_hindcast,
              method = "loess",
              aes(color = model), span = 30/n, linewidth = 0.5, se = FALSE, n = n) +
  scale_x_datetime(NULL, date_breaks = "1 month", 
                   date_labels = "%b", expand = c(0, 0)) +
  scale_y_continuous(NULL, labels = \(x) labels_extent(x, units = "M km²/day")) +
  scale_color_models +
  coord_cartesian(xlim = as.POSIXct(c("2000-01-01", "2000-12-31"), tz = "UTC"))
```

@@

ice_shelves_nsicd <- fread(here::here("data/derived/antarctic_ice_shelves_nsidc.csv"))

hindcast_clim |> 
  # dcast(time + xgrid + ygrid ~ model, value.var = "aice") |> 
  # setnames(which_dataset, "obs") |>
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = aice - obs, fill = after_stat(level)), 
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised(paste0("Bias with respect to ",
                                          labels_models[which_dataset]),
                                   low = scales::muted("red"),
                                   high = scales::muted("blue")) +
  # geom_polygon(data = ice_shelves_nsicd, aes(X, Y, group = group), 
  #              colour = "black", fill = "#FAFAFA") +
  geomcoord_antarctica +
  geom_antarctica_fill +
  facet_wrap(~ month) +
  wide_legend +
  tag_facets()
```

@@

clim_std2 |> 
  rbind(errors) |>
  _[measure == "rmse"] |>
  _[, time2 := update(time, year = 2000)] |> 
  # merge(clim_std) |>
  _[obs_dataset ==  which_dataset] |>
  _[, median_ci(value),
    by = .(lag, version, measure, month(time_forecast), obs_dataset)] |>
  ggplot(aes(lag, mid)) +

@@

  # geom_ribbon(aes(ymin = low, 
  #                 ymax = high, 
  #                 color = version, 
  #                 fill = version,
  #                 group = interaction(obs_dataset, version)),
  #             alpha = 0.1) +
  geom_line(aes(color = version, group = interaction(obs_dataset, version)), linewidth = 1) +

  # geom_line(aes(y = clim_std)) +
  scale_x_continuous(breaks = NULL, expand = c(0, 0)) +
  labs(y = "RMSE", 
       x = "Lag") +
  scale_color_models +
  scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month)) 
```

@@

@@

```{r}
topo_lonlat <- rcdo::cdo_topo(grid = "r360x180") |>
  rcdo::cdo_execute(options = c("-f nc")) |>
  metR::ReadNetCDF(c(z = "topo"), subset = list(lat = c(-90, -60))) |> 
  ggperiodic::periodic(lon = c(0, 360))

topo_lonlat <- here::here("data/raw/ETOPO.nc") |> 
  ReadNetCDF() |> 
  na.omit() 
```

@@

binwidth <- 0.015

rmse_lon_mean |> 
  _[obs_dataset == which_dataset] |> 
  ggperiodic::periodic(lon = c(0, 360)) |> 
  ggplot(aes(lon, lag)) +
  geom_contour_fill(aes(z = dif, fill = after_stat(level)),
                    breaks = AnchorBreaks(0, exclude = 0, binwidth = binwidth))  +
  geom_contour_tanaka(aes(z = dif), range = c(0.01, 0.4),
                      breaks = AnchorBreaks(0, exclude = 0, binwidth = binwidth)) +

  geom_contour_fill(data = topo_lonlat,
                    aes(lon, scales::rescale(lat, c(-60, 1)), 
                        z = z, 
                        fill = NULL),
                    fill = "#FAFAFA",
                    breaks = c(0, Inf)) +

  geom_contour2(data = topo_lonlat,
                aes(lon, scales::rescale(lat, c(-60, 1)), z = z),
                breaks = 0) +
  geom_hline(data = month_from_lag(1, 1), aes(yintercept = lag),
             colour = "gray50", linewidth = 0.1) +

@@

  scale_y_continuous(breaks = NULL, expand = c(0, 0)) +
  labs(fill = NULL, 
       y = "Lag") +
  wide_legend +
  facet_wrap(~ month, labeller = labeller(month = labels_month)) +
  theme(panel.background = element_blank(), panel.grid = element_blank(),
        tagger.panel.tag.background = element_rect(colour = "white", fill = "white")) +
  tag_facets(position = list(x = 0.08, y = 1, hjust = 0, vjust = 1))
```
